---
title: "Homework 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Masked objects

Load `dplyr`.

```{r, warning = FALSE, echo = TRUE}
library(dplyr)
```

Does the namespace of `dplyr` overalap with namespaces that you already
have loaded?  For which packages and functions?

```{r}
# The `filter` and `lag` functions from the `stats` package and the
# `intersect`, `setdiff`, `setequal`, and `union` functions from `base`
# are masked because those names are also used in the `dplyr` namespace.
```

Look at `?intersect` and pull up the help pages for the base version and
the dplyr version.  If you want to use the base version but you have
`dplyr` loaded, will it cause a problem?

```{r}
# It looks like it would probably be okay.  Although the `dplyr` version of
# `intersect` would get used if you didn't specify which version you wanted,
# the default method of the `dplyr` version is to call the `base` version.
```

Look at `?filter` and pull up the help pages for the `stats` version and
the `dplyr` version.  If you want to use the base version but you have
`dplyr` loaded, will it cause a problem?

```{r}
# In this case it probably would, since the two filter functions take different
# arguments and do different things.
```

With `dplyr` loaded, how would you call the non-`dplyr` version of one of
these functions?

```{r eval = FALSE}
# Example for `filter`
stats::filter(1:100, c(1,1,1))
```

