---
title: "Lab 7"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## WorldClim data

WorldClim (www.worldclim.org) provides current, historical, and future predicted
climate data for use in ecological and GIS studies.  I downloaded several files
from that site for you to work with today.

```{r echo = TRUE}
# Average June temperature 1970-2000, deg. C
juneTempsFile <- "wc2.0_10m_tavg_06.tif"
# Average June temperature 1960-1990, deg. C * 10
juneTempsFile2 <- "tmean6.bil"
# Average annual temperature 1970-2000, deg. C
avgTempFile <- "wc2.0_bio_10m_01.tif"
```

The packages "raster" and "rgdal" can be used for importing and manipulating this
sort of data.

```{r eval = FALSE, echo = TRUE}
install.packages("raster")
install.packages("rgdal")
```

```{r, warning = FALSE, echo = TRUE}
library(raster)
library(rgdal)
```

We can preview the data with `plot`.

```{r, echo = TRUE, warning = FALSE}
library(viridis)
plot(raster(juneTempsFile), col = plasma(255))
```

## Number formats in the WorldClim datasets

For each of the three file names, use the function `GDALinfo` to learn more
about the file's contents.  Pay particular attention to the format listed
under "GDtype".

**Question 1 to turn in:**  What does the number after "Int" or "Float" mean?
Which file(s) contain numbers in a format that matches one of the formats in
the R base?  Given the number format of "tmean6.bil", why do you think that
file contains degrees Celsius multiplied by ten, while the other files just
contain degrees Celsius without a multiplier?

```{r eval = FALSE}
# The number indicates how many bits the format uses.  "wc2.0_bio_10m_01.tif"
# uses 64-bit floats (double-precision floating point numbers), which is the
# 'numeric' class and 'double' type in the R base.  The other two files use
# formats not found in the R base; 32-bit floats (single-precision floating
# point) for "wc2.0_10m_tavg_06.tif" and 16-bit integers for
# "tmean6.bil".  Because "tmean6.bil" stores temperatures in an integer format,
# by multipling by ten it allows temperatures to be stored precisely in 
# 1/10 degree increments.

GDALinfo(juneTempsFile)
GDALinfo(juneTempsFile2)
GDALinfo(avgTempFile)
```

## Integer math

Let's extract the average June temperature from 1960-1990 for the collection
locations in our *Miscanthus sinensis* data set from Week 1.

```{r echo = TRUE}
msiyield <- read.csv("Miscanthus_sinensis_yield.csv")
msiXY <- as.matrix(msiyield[,3:2])
colnames(msiXY) <- c("x", "y")
# remove rows without spatial coordinates
msiXY <- msiXY[!is.na(msiXY[,1]), ]

# make raster object for tmean6.bil
juneTempsRaster2 <- raster(juneTempsFile2)
# extract cell numbers
msiCells <- cellFromXY(juneTempsRaster2, msiXY)

# get temperatures
msiJuneTemp <- getValues(juneTempsRaster2)[msiCells]
msiJuneTemp
typeof(msiJuneTemp)
```

Say you want to convert degrees to growing degrees, so that you can calculate
growing degree days.  Typically you subtract 10°C from a 
temperature in order to convert to growing degrees (GD), since many crops will 
not grow below 10°C.  Temperatures below 10°C are also converted to 0 GD, and 
temperatures above 30°C are capped at 20 GD since most crops do not have 
increased productivity above that temperature.

* 5°C --> 0 GD
* 10°C --> 0 GD
* 15°C --> 5 GD
* 30°C --> 20 GD
* 35°C --> 20 GD
* etc.

The values that we got from the WorldClim file are integers, indicating 
temperature times ten.  Pretend that we want to keep them that way in order to 
save memory and processing time.

**Question 2 to turn in:** Make a function to convert degrees Celsius to 
growing degrees.  The output should still be in integer format and still
multiplied by ten, like the input.

```{r}
# function to take temperature, in degrees C times 10, and convert it to
# growing degrees, also in degrees C times 10, in integer format.
growingDegreesTimesTen <- function(temp){
  gd <- temp - 100L
  gd[gd < 0L] <- 0L
  gd[gd > 200L] <- 200L
  return(gd)
}
```

```{r echo = TRUE}
myGD <- growingDegreesTimesTen(msiJuneTemp)
myGD[1:15]
typeof(myGD)
```


**Hints**

You should be able to keep the numbers in integer format without using 
`as.integer`.

You can use Boolean indexing to identify and replace values below zero or
above 20.

Remember that since everything is mutliplied by 10, you actually are 
subtracting 100 rather than 10, and capping at 200 rather than 20.

## Package-specific classes

We used some specialized functions to import and examine GIS data.  The object
storing the data is a specialized class for that type of data.  In addition to
including function definitions, the `raster` package includes class definitions.

In lecture we talked about how `plot` is a generic function.  It must have a 
method for whatever class we have our GIS data in, since we were able to
use `plot` directly on that object.

Try to find the help page for the class, as well as the help page for the `plot`
method.

**Question 3 to turn in:** What class is `juneTempsRaster2`?  What code pulls
up the help page for the `plot` method?

```{r eval = FALSE}
## The class of `juneTempsRaster2` is `"RasterLayer"`.
class(juneTempsRaster2)

## To get the help page for the `plot` method:
?`plot,Raster,ANY-method`
# or
?`plot,Raster,Raster-method`
```


**Hints**

The class name itself has a "package" attribute indicating what package the
class came from.  Be sure not to confuse the class name and package name.

The class is a particular type of class called S4, which is a system for
package developers to add new classes.  So far in lecture we have mostly looked
at S3 classes and methods, so the documentation might be slightly different
from what you are used to.

To find the help pages, you might begin typing and then look at the 
auto-complete suggestions.