---
title: 'R lab 4: Graphing with ggplot2'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages for today

Load the following packages.  Use `install.packages` if they aren't on your
computer already.

```{r}
library(viridis) # this has some nice color-blind-friendly color gradients
library(ggplot2) # this will make plotting much easier
```

Also read in our *Miscanthus sinensis* dataset.

```{r}
msi_data <- read.csv("Miscanthus_sinensis_yield.csv", row.names = 1)
```

## Intro

A week ago we made some plots of our data using the R base graphics system.
Plots made that way are highly customizable, just not in a way that is very
user-friendly.  For example, here is what I would do if I wanted to plot
biomass yield versus plant height and then color by stem diameter:

```{r}
minDiam <- min(msi_data$Stem.diameter, na.rm = TRUE)
maxDiam <- max(msi_data$Stem.diameter, na.rm = TRUE)
stemDiamIndex <- round((msi_data$Stem.diameter - minDiam)/
                         (maxDiam - minDiam) * 100) + 1
plot(msi_data$Plant.height, msi_data$Biomass.yield,
     col = viridis(101)[stemDiamIndex],
     xlab = "Plant height", ylab = "Biomass yield")
```

And I don't even get a legend for what the colors mean.  I would have to 
build that manually.

Here is the same plot using the package `ggplot2`

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield,
                 color = Stem.diameter)) +
  scale_color_viridis()
```

So, for this lesson we'll continue using `ggplot2` since we can cover a lot
more ground in terms of different kinds of plots!

## Syntax of ggplot2

When you create a plot with `ggplot2`, the first part of the expression is
always a call to the function `ggplot`.  Typically the only argument you
give `ggplot` is the data frame with the data that you want to plot.  The
call to `ggplot` is always followed by a `+` sign.

Next, we add a *geometry* to the expression.  There are a bunch of functions
in `ggplot2` that start with `geom` and can go here.  In the lower right part
of your RStudio screen, click "Packages", then "ggplot2" to see a list of the
different geometries that are available.

Either within the call to `ggplot` or within the geometry function, we need to
specify the *aesthetics* using the `aes` function.  This is where we specify
how our variables should correspond to different aspects of the plot.

There are a few more things that we can add to the expression after that,
and we'll get into them below.

## More about aesthetics

Notice that every argument we pass to `aes` is a column name from our data
frame.  We used the arguments `x`, `y`, and `color`, but more are available,
and can be found at `?geom_point`.  Say we wanted shapes for the different
genetic groups.  We can only get shapes for six different groups, so we'll
eliminate some.

```{r}
ggplot(msi_data[msi_data$Genetic.group %in% c("S Japan", "N Japan", "SE China/tropical",
                                              "Yangtze-Qinling", "Sichuan", 
                                              "Korea/N China"), ]) +
  geom_point(aes(x = Plant.height, y = Biomass.yield,
                 color = Stem.diameter, shape = Genetic.group)) +
  scale_color_viridis()
```

We had color for stem diameter, but how about size for number of stems?

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield,
                 color = Stem.diameter, size = Number.of.stems)) +
  scale_color_viridis()
```

Hmm, it might be more readable if there were a black line around each circle.
In R that's a different shape.  If you Google "R pch" you will see a 
set of numbered shapes that can be used for plotting.  21 through 25 are
shapes with outline and fill colors.  To set an aesthetic for the whole plot,
rather than having it correspond to a variable, we will put it within the
call to `geom_point` rather than the call to `aes`.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield,
                 color = Stem.diameter, size = Number.of.stems),
             shape = 21) +
  scale_color_viridis()
```

Whoops, this caused stem diameter to control the outline color.  We can use
`fill` instead to control the fill color.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield,
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21) +
  scale_fill_viridis()
```

To make the circles semi-transparent, we can adjust the `alpha` aesthetic.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield,
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis()
```

So, we can see that the tallest and highest-yielding plants didn't have the most
stems, but they did tend to have thick stems.

**Exercise 1**

Try having one variable control the shape outline color, and another control the
fill color.

```{r echo = FALSE, eval = FALSE}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield,
                 fill = Stem.diameter, color = Number.of.stems),
             shape = 21, alpha = 0.5, size = 2) +
  scale_fill_viridis()
```

### Transforming variables

There are a lot of low-yielding plants and just a few very high yielding plants.
Yield also results from the plant's size in three-dimensions, whereas height and
stem diameter are one-dimensional measurements.  So maybe it would be easier to 
visualize the cube root of yield rather than yield itself.  We can do that 
transformation right in the call to `aes`.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis()
```

## Facets

There's still a lot of information in that plot, and we can't see everything.
Maybe we want to split it up by genetic group.  We can do that by adding 
*facets*, here with a function called `facet_wrap`.  The `~` sign is needed,
because a formula is used to generate the facets, even if it is just one
variable.

```{r}
ggplot(msi_data[msi_data$Genetic.group %in% c("S Japan", "N Japan", "SE China/tropical",
                                              "Yangtze-Qinling", "Sichuan", 
                                              "Korea/N China"), ]) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis() +
  facet_wrap(~ Genetic.group)
```

We can also make facets based on a Boolean vector.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis() +
  facet_wrap(~ Number.of.stems < 250)
```

What if we want to facet using two variables?  Then use `facet_grid`.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis() +
  facet_grid(Stem.diameter < 6 ~ Number.of.stems < 250)
```

**Exercise 2**

Make facets based on the latitude and longitude of origin of the plants.

```{r echo = FALSE, eval = FALSE}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis() +
  facet_grid(Latitude > 0 ~ Longitude > 0)
```

## Additional plot geometries

If you look at the list of functions for `ggplot2`, (available under the "Packages"
tab in the lower right), you will find many `geom` functions in addition to 
`geom_points`.  They can actually be layered on top of each other.  For example,
we can add a curve showing the overall trend with `geom_smooth`.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis() +
  geom_smooth(aes(x = Plant.height, y = Biomass.yield ^ (1/3)))
```

This can be forced to be a trendline made from linear modeling:

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis() +
  geom_smooth(aes(x = Plant.height, y = Biomass.yield ^ (1/3)),
              method = "lm")
```

An alternative to `geom_points` for really dense scatter plots is `geom_bin2d`.

```{r}
ggplot(msi_data) +
  geom_bin2d(aes(x = Plant.height, y = Biomass.yield ^ (1/3)),
             bins = 50)
```

Or `geom_density2d`:

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 size = Number.of.stems),
             shape = 16, alpha = 0.5) +
  geom_density2d(aes(x = Plant.height, y = Biomass.yield ^ (1/3)),
                 color = "green")
```

If you want to see continuous distributions by groups, you can make a 
violin plot with `geom_violin`.

```{r}
ggplot(msi_data) +
  geom_violin(aes(x = Genetic.group, y = Plant.height))
```

To rotate it and make the labels more readable, I can use `coord_flip`.

```{r}
ggplot(msi_data) +
  geom_violin(aes(x = Genetic.group, y = Plant.height)) +
  coord_flip()
```

**Exercise 3**

Make a violin plot of plant height, divided into two groups by stem diameter.

```{r echo = FALSE, eval = FALSE}
ggplot(msi_data) +
  geom_violin(aes(x = Stem.diameter > 6, y = Plant.height))
```

```{r echo = FALSE, eval = FALSE}
ggplot(msi_data) +
  geom_violin(aes(x = cut_number(Stem.diameter, n = 4), y = Plant.height))
```

## Colors

We have already used colors a little bit.  The default color scheme is a scale
of light to dark blue, which we changed to a more visually dynamic scale using
`scale_color_viridis` and `scale_fill_viridis`.  There are five color scales
in the `viridis` package, called "magma", "inferno", "plasma", "viridis",
and "cividis".  All are designed to be color-blind friendly.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_viridis(option = "plasma")
```

If you need colors for discrete data, like our genetic groups, there are some
good ones provided by Color Brewer.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Genetic.group),
             shape = 21, alpha = 0.8, size = 2) +
  scale_fill_brewer(type = "qual", palette = "Paired")
```

Here is a good cheat sheet for making colors in R:

https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf

There are lots of named colors in R, as listed on that sheet.  We can use
those named colors within the geometry function if we want everything to be
the same color.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3)),
             shape = 16, alpha = 0.5, size = 2, color = "magenta")
```

You can also build your own color schemes.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Genetic.group),
             shape = 21, alpha = 0.8, size = 2) +
  scale_fill_manual(values = c("SE China/tropical" = "purple",
                    "Sichuan" = "orange",
                    "Yangtze-Qinling" = "darkgreen",
                    "Ornamental" = "yellow3",
                    "US naturalized" = "yellow4",
                    "N Japan" = "blue",
                    "S Japan" = "yellow",
                    "Korea/N China" = "red",
                    "M. sacchariflorus 2x" = "cyan",
                    "M. sacchariflorus 4x" = "cyan",
                    "M. xgiganteus 2x" = "cyan3",
                    "M. xgiganteus 3x" = "cyan3"))
```

And your own color gradients.

```{r}
ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_gradient(low = "turquoise", high = "tomato")
```

What if you need to match a particular color from another piece
of software?  If you have the RGB values, you can make a string that
will give you that color.

```{r}
rgb(100, 30, 210, maxColorValue = 255)
```

Or you can do the opposite, if you want to recreate a color in other
software that you used in R.

```{r}
col2rgb("turquoise")
```

**Exercise 4**

Generate a character string representing colors with the RGB values
200,0,40 and 0,220,200, respectively.  Make a scale between these
two colors using `scale_fill_gradient`.  Use that gradient in a plot.

```{r echo = FALSE, eval = FALSE}
col1 <- rgb(200, 0, 40, maxColorValue = 255)
col2 <- rgb(0, 220, 200, maxColorValue = 255)

ggplot(msi_data) +
  geom_point(aes(x = Plant.height, y = Biomass.yield ^ (1/3),
                 fill = Stem.diameter, size = Number.of.stems),
             shape = 21, alpha = 0.5) +
  scale_fill_gradient(low = col1, high = col2)
```

