---
title: "Lab 10 - Making an Rcpp function"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE, eval = FALSE}
## Generate some data for this lab
# recombination frequencies
recom <- rbeta(499, 2, 10)/2
# matrix for SNP genotypes
TestCrossMat <- matrix(0L, nrow = 300, ncol = 500,
                       dimnames = list(paste("Progeny", 1:300),
                                       paste("Marker", 1:500)))
for(p in 1:nrow(TestCrossMat)){
  for(m in 1:ncol(TestCrossMat)){
    if(m == 1){
      TestCrossMat[p,m] <- sample(c(0L, 1L), 1)
    } else {
      TestCrossMat[p,m] <- ifelse(runif(1) <= recom[m-1], 1L - TestCrossMat[p,m-1], TestCrossMat[p,m-1])
    }
  }
}
save(TestCrossMat,  file = "TestCrossMat.RData")
```

## Finding recombination break points

Say we have a matrix of SNP data from a testcross mapping population, with all
homozygotes coded as `0` and heterozygotes coded as `1`.  The markers are in
map order.

```{r echo = TRUE}
load("TestCrossMat.RData")
TestCrossMat[1:10,1:10]
```

We want to find recombination breakpoints, i.e. locations of markers
where the genotype is different from the genotype at the adjacent marker. 
Here's a loop to do this somewhat inefficiently, although some allocation
is done.  The output vector containing the indices is
preallocated to a longer length than needed, then shortened when the
loop is done.

```{r}
# A function that looks for breakpoints in just one progeny.
findBreakpointsOneProgeny <- function(genotypes){
  nloc <- length(genotypes) # number of markers
  # preallocate a vector to store the breakpoints.  bigger than needed.
  outindices <- integer(nloc)
  # variable to count up how many indices we have found
  mycount <- 0L
  
  # loop through markers and look for breakpoints
  for(m in 2:nloc){
    if(genotypes[m] != genotypes[m - 1]){
      mycount <- mycount + 1L
      outindices[mycount] <- m
    }
  }
  
  # trim vector down to right size
  if(mycount > 0){
    outindices <- outindices[1:mycount]
  } else {
    outindices <- integer(0)
  }
  
  return(outindices)
}

# apply the function across the matrix
mybp <- apply(TestCrossMat, 1, findBreakpointsOneProgeny)
mybp[1:5]
```

Did `findBreakpointsOneProgeny` get automatically compiled?

## Translating the function to C++

In RStudio, make a new C++ file using the "New File" dialogue.  As you edit the
file, be sure to keep a header like this (with a comment to describe the function):

```{Rcpp echo = TRUE, eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// Insert function description here
// [[Rcpp::export]]
```

Take the above function for finding breakpoints (`findBreakpointsOneProgeny`) 
and translate it into C++.  A few things to remember:

* Declare your variables and put a semicolon at the end of each command.
* You can change the value of a variable as many times as you want, but only need to declare
its class once (generally at the beginning of the function).
* The first item in a vector is index `0`, the second item index `1`, and so on.
* `i++;` is shorthand for `i = i + 1;`
* To preallocate an integer vector of length 5, do `IntegerVector myvect(5);`,
where `myvect` is the name of the vector.
* You can find the length of a vector `myvect` with `myvect.size()`.
* `Range(0, 3)` in Rcpp is the equivalent of `0:3` in R, and can be used for 
indexing vectors.
* You can reset a vector to have length zero like this: `myvect = IntegerVector(0);`
* You don't need parentheses after your `return` statement.

```{Rcpp echo = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// Function to find recombination breakpoints in a vector of SNP genotypes.
// Genotype values must be 0 or 1.

// [[Rcpp::export]]
IntegerVector findBreakpointsOneProgenyRcpp(IntegerVector genotypes){
  int nloc = genotypes.size();    // Number of markers
  IntegerVector outindices(nloc); // preallocate vector for output
  int mycount = 0;                // count the number of indices found
  
  // loop through markers and look for breakpoints
  for(int m = 1; m < nloc; m++){
    if(genotypes[m] != genotypes[m - 1]){
      outindices[mycount] = m + 1;
      mycount++;
    }
  }
  
  // Trim down output vector
  if(mycount > 0){
    outindices = outindices[Range(0, mycount - 1)];
  } else {
    outindices = IntegerVector(0);
  }
  
  return outindices;
}
```

```{r echo = TRUE}
# example use of compiled function
mybp2 <- apply(TestCrossMat, 1, findBreakpointsOneProgenyRcpp)
mybp2[1:5]
identical(mybp, mybp2)
```

**Question 1 to turn in (2 points):** Turn in the C++ code for your translation of the 
`findBreakpointsOneProgeny` function.

## Comparing R vs. C++ for this function

Use `microbenchmark` to run a test comparing the speed of the function written 
in R to the version you translated into C++.  

**Question 2 to turn in (1 point):** Which function is faster, or are they
about the same?  Which would you rather include in an R package and why?

```{r echo = FALSE, eval = FALSE}
library(microbenchmark)
microbenchmark(apply(TestCrossMat, 1, findBreakpointsOneProgeny),
               apply(TestCrossMat, 1, findBreakpointsOneProgenyRcpp))

# Although the R version got compiled, the C++ version is still ~50x faster.
# In this case it is probably worth it to use the Rcpp version in an R 
# package.  It will make things a little more complicated for someone looking
# at the package source, but much faster for end users.
```

## Homework, or if you have extra time in lab

Even with the Rcpp function, we're still `apply`ing it across the matrix,
which could slow things down a little bit.  Try writing code in R that
processes the whole matrix at once, reducing what happens inside loops 
or `apply` calls.  Use `microbenchmark` or profiling to compare it to
what we did above.

```{r echo = FALSE}
findBreakpointsAll <- function(genmat){
  nloc <- ncol(genmat)
  # get true/false matrix for genotypes being identical to adjacent gen.
  idmat <- genmat[,1:(nloc-1)] != genmat[,2:nloc]
  # apply `which` across rows to get indices for all
  outind <- apply(idmat, 1, function(x) unname(which(x) + 1L))
  return(outind)
}
```

```{r echo = TRUE}
mybp3 <- findBreakpointsAll(TestCrossMat)
identical(mybp, mybp3)
```
```{r eval = FALSE, echo = TRUE}
microbenchmark(findBreakpointsAll(TestCrossMat),
               apply(TestCrossMat, 1, findBreakpointsOneProgeny),
               apply(TestCrossMat, 1, findBreakpointsOneProgenyRcpp))
```

