---
title: "Homework 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
```

## Dataset for this homework

These homeworks are not graded, but are provided to give you extra practice.

Let's explore the `airquality` example dataset that comes with R.

```{r echo = TRUE}
data(airquality)
```

Remember that you can learn more about it by reading its help page:

```{r echo = TRUE, eval = FALSE}
?airquality
```

## Exercises

(1) Use some of the functions that you have learned for previewing the data frame.
How many rows and columns does it have?  What is the data type of each column?
What is the name of each column?

```{r}
head(airquality)
str(airquality)
```

(2) Sort the data frame by temperature, then by wind speed.

```{r}
# in the R base:
airquality[order(airquality$Temp, airquality$Wind),]

# in dplyr
library(dplyr)

airquality %>%
  arrange(Temp, Wind)

## Note that both of those commands display the sorted data, but don't change
## the original data frame.
```

(3) Make a new data frame from the original, containing only the observations
from August and September.  Omit the Wind and Temp columns from this new data frame.

```{r}
# in the R base
# (Your code might look slightly different but accomplish the same thing.)
airquality_sub <- airquality[airquality$Month %in% c(8,9),
                             c("Ozone", "Solar.R", "Month", "Day")]

# in dplyr
airquality_sub <- airquality %>%
  filter(Month %in% c(8,9)) %>%
  select(Ozone, Solar.R, Month, Day)
```

(4) Make a scatter plot of ozone vs. wind speed.

```{r}
# in the R base
plot(airquality$Wind, airquality$Ozone)
```

```{r}
# in ggplot2
library(ggplot2)
ggplot(airquality, aes(x = Wind, y = Ozone)) +
  geom_point()
```

(5) Within each month, get the mean of the first four columns (Ozone, Solar.R, 
Wind, and Temp).  Since there are some missing data, you should set the 
`na.rm` argument of `mean` to `TRUE`.

```{r}
# in the R base; there is one function called `by` that we haven't covered yet,
# that works like tapply but on data frames.
by(airquality[,1:4], airquality$Month, colMeans, na.rm = TRUE)

# or we could use tapply on individual columns
tapply(airquality$Ozone, airquality$Month, mean, na.rm = TRUE)
tapply(airquality$Solar.R, airquality$Month, mean, na.rm = TRUE)
tapply(airquality$Wind, airquality$Month, mean, na.rm = TRUE)
tapply(airquality$Temp, airquality$Month, mean, na.rm = TRUE)

# in dplyr:
airquality %>%
  group_by(Month) %>%
  summarize(mean_ozone = mean(Ozone, na.rm = TRUE),
            mean_solar.r = mean(Solar.R, na.rm = TRUE),
            mean_wind = mean(Wind, na.rm = TRUE),
            mean_temp = mean(Temp, na.rm = TRUE))
```

(6) Extract a vector of ozone values, but only for days where the solar
radiation was below 100 in the month of May.

```{r}
# Use `which` rather than direct Boolean indexing, because there are some
# missing data.
airquality$Ozone[which(airquality$Solar.R < 100 & airquality$Month == 5)]

# with dplyr:
filter(airquality, Solar.R < 100, Month == 5)$Ozone
```

(7) Say you want to take the mean of a vector called `myvect`, and you want to
ignore `NA` (missing) values.  Which of the following function calls will work?
Which won't?  Why?

```{r echo = TRUE, eval = FALSE}
mean(myvect, TRUE)             # 1
mean(myvect, na.rm = TRUE)     # 2
mean(na.rm = TRUE, x = myvect) # 3
mean(myvect, 0, TRUE)          # 4
mean(x = myvect, na.rm = TRUE) # 5
```

```{r}
# 1 won't work; `TRUE` gets passed to `trim`, the second argument.
# 2 will work; myvect gets passed to x as the first argument, and TRUE is
# passed to the correct named argument.
# 3 will work; both arguments are named, so the order doesn't matter.
# 4 will work; arguments are not named but are in the correct order.
# 5 will work, like # 3.
```

(8) Say you want to sort `airquality` by Day, and so you write the following
code:

```{r echo = TRUE}
airquality %>%
  arrange(Day)
```

However, when you run

```{r echo = TRUE}
head(airquality)
```

you see that the data frame is still in the original order.

Rewrite the code to permanently sort the data frame for future use.

```{r}
airquality <- airquality %>%
  arrange(Day)
```

