---
title: "Homework 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In lecture and lab we had some examples using real datasets.  For this homework
I'm providing some more abstract examples so that you can think about how the code
is working and how to write it.

## IF/ELSE syntax

**1.** The following code gives an error.  How would you fix it?

```{r echo = TRUE, error = TRUE}
x <- 3
if(x < 5){
  print("Less than five")
}
else{
  print("At least five")
}
```

```{r eval = FALSE}
# By putting `else` on the same line as the closing curly brace, 
# we show R that the command is not done yet.
x <- 3
if(x < 5){
  print("Less than five")
} else { 
  print("At least five")
}
```

**2.** The following code works but is not well formatted.  Fix the formatting
while retaining the same functionality.

```{r echo = TRUE}
x <- 3
if(x > 7)
print("Greater than seven")
print("Probably a number")
```

```{r eval = FALSE}
# probably the most readable:
if(x > 7){
  print("Greater than seven")
}
print("Probably a number")

# also okay if you prefer it this way:
if(x > 7) print("Greater than seven")

print("Probably a number")

# or
if(x > 7)
  print("Greater than seven")

print("Probably a number")

# Note that the last two only work for cases where there is only one line
# under control of the IF statement.
```

## Loops

**3.** We can use the `sample` function to draw a random integer.  For example, this
command will get us one random integer from 1 to 1000.

```{r echo = TRUE}
sample(1000, 1)
```

Write a loop to do the following:  Draw a random integer.  If the integer is
equal to 500, stop the loop.  Otherwise, try again.  Save the integer that is
closest to 500 (the `abs` function might be helpful).  If 200 attempts have
been made, stop the loop.  You can do this with `for`, `while`, or even `repeat`,
depending on your preference.  After the loop has run, you should have a variable
containing the closest integer to 500 that you found, and a variable indicating
how many attempts were made.

```{r}
## example with a while loop ##

# dummy value that will definitely be further than anything we will sample
closest <- 10000 
# variable to count up how many attempts we have made
attempts <- 0
while(closest != 500){
  newint <- sample(1000, 1)
  if(abs(newint - 500) < abs(closest - 500)){
    closest <- newint
  }
  attempts <- attempts + 1
  if(attempts >= 200) break
}

## example with a for loop ##
closest <- 10000
for(attempts in 1:200){
  newint <- sample(1000, 1)
  if(abs(newint - 500) < abs(closest - 500)){
    closest <- newint
  }
  if(closest == 500) break
}

## example with repeat ##
closest <- 10000
attempts <- 0
repeat{
  newint <- sample(1000, 1)
  if(abs(newint - 500) < abs(closest - 500)){
    closest <- newint
  }
  if(closest == 500) break
  attempts <- attempts + 1
  if(attempts >= 200) break
}
```

**4.** The following code loops through a vector, but attempts to make the vector
shorter at the same time, eliminating the last element on each iteration.

```{r echo = TRUE, eval = FALSE}
myvect <- 1:20
for(i in myvect){
  cat(i, sep = "\n")
  myvect <- myvect[-length(myvect)]
}
```

What happens when you run this code?  What does `myvect` look like at the
end?  How far through the vector did `i` get?  What does this tell you 
about passing a vector to a `for` statement?

```{r}
# The numbers 1 through 20 are printed out.  `myvect` was reduced to an empty
# vector because all 20 elements were removed, one at a time.  However, `i`
# got all the way to 20.  When we passed `myvect` to `for`, a copy of `myvect`
# was made for iteration, and was not affected by anything we did to the 
# original.
```

**5.** Say you wanted to make a loop that mostly worked like a `for` loop,
moving forward through a vector, but under certain conditions would move
backwards through the vector.  How might you accomplish this?

```{r eval = FALSE}
# some vector that we want to loop through; let's take 100 random numbers
myvect <- runif(100)
# variable to hold the index
i <- 1
# instead of a for loop, make a while loop to check whether we have reached the end
while(i <= length(myvect)){
  print(myvect[i])
  # increment i under some conditions, and decrement under others.
  # for this example, we'll randomly decide.
  rand <- runif(1)
  if(rand < 0.95){
    i <- i + 1
  } else {
    i <- i - 1
  }
}
```

## Functions

**6.** Look at the following script:

```{r echo = TRUE}
miles_to_km <- function(dist){
  dist <- dist * 1.60934
  return(dist)
}

dist <- 5
dist_km <- miles_to_km(dist = dist)
```

Is this going to cause any problems in terms of variable scope?

```{r}
# No, this is fine, and in fact it is often convenient to have variables
# with the same names in different scopes if they represent the same thing.
# Because there is an argument called dist, the miles_to_km will use that
# dist and not bother searching for one in the global environment.
```

**7.** Can you define a function within the body of another function?
What variable scopes would that function have access to?

```{r eval = TRUE}
# Yes. Here's an example.  The function has access to its own scope,
# the scope of its containing function, and the global environment.

# another version of our function for calculating stem volume
stem_vol <- function(height, diam){
  crossection_area <- function(diam){
    return(pi * (diam/2) ^ 2)
  }
  
  return(height * crossection_area(diam))
}

stem_vol(200, 0.4)

# note this version will also work, although maybe it is not the best style
stem_vol <- function(height, diam){
  crossection_area <- function(){
    return(pi * (diam/2) ^ 2)
  }
  
  return(height * crossection_area())
}

stem_vol(200, 0.4)

# Functions-within-functions can be especially handy when you are building
# a long complicated function, and need to break the algorithm up into 
# smaller chunks, but don't expect to use those chunks anywhere else.
```

